Phenotypic evolution is more constrained in simpler food webs

### 2018-11-22 manuscript.Rmd
incorporated Jordi's comments on Results.

**Selection surface**: Since larval survival was our measure of absolute fitness, we used a GLMM that assumed a binomial error distribution (and logit-link function). To approximate the selection surface, we modelled larval survival as a function of food-web treatment as well as linear ($\alpha_{z_i}$), quadratic ($\alpha_{z_i:z_i}$), and linear interactions ($\alpha_{z_i:z_j}$) between each trait. We also allowed these trait-fitness relationships ($\alpha$) to vary between food-web treatments. Note that to obtain valid estimates of linear trait-fitness relationships, we removed nonlinear terms prior to estimating linear relationships [@Lande1983]. Other approaches have been advocated for approximating selection surfaces [@Schluter1988]; however, our approach enables us to calculate selection gradients, and thus is more appropriate for approximating the adaptive landscape [@Arnold2003]. To account for the nonindependence of clutch size (measured at gall level) and oviposition preference (measured at plant level) as well as any independent effects of willow genotype on larval survival, we modelled gall ID nested within plant ID nested within genotype ID as random effects. Although statistical models with random effects are not common in analyses of natural selection, we think this is a mistake since modelling random effects can mitigate biased estimates of selection due to environmental covariances between traits and fitness [@Rausher1992]. Since our end goal was to characterize the relationship between mean trait values and mean fitness (adaptive landscape), we assumed the mean value of our random effects (i.e. setting them to zero) when calculating trait-fitness relationships. We then used parametric bootstrapping (1,000 replicates) to estimate the effect of food-web treatment on larval survival as well as trait-fitness relationships in each food-web treatment. To determine whether trait-fitness relationships differed between food-web treatments, we calculated the difference in bootstrapped replicates between treatments.

@@

**Adaptive landscape**: We took advantage of existing theory that translates estimates of selection gradients to the adaptive landscape [@Phillips1989; @Arnold2003]. The shape of the adaptive landscape can be described by both its slope and curvature, each of which describe different aspects of constraints on phenotypic evolution. Specifically, the slope corresponds to the column vector of directional selection gradients, which in our study corresponds to:

@@

$$\text{Slope} = \beta = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} $$
This slope describes the selective constraints acting on the trajectory of phenotypic evolution, which is governed by the 'Lande equation', $\Delta \bar z=\text{G}\beta$ [@Lande1979]. Quantitative predictions of evolutionary trajectories requires knowledge of a population's **G**-matrix; however, we can gain qualitative insight to evolutionary constraints by counting the number of phenotypic traits under directional selection [selective constraints, @Arnold1992].

The curvature (**C**) of the adaptive landscape is a matrix that is determined by directional, quadratic, and correlational selection gradients, such that \boldmath$\text{C}=\gamma - \beta \beta^\text{T}$, or more explicitly:

@@

Note that we ommitted the upper triangle of each matrix for clarity since it is simply the reflection of the lower triangle. This curvature matrix describes the indirect effects of selection on genetic constraints within a generation ($\Delta \text{G}=\text{G}(\gamma -\beta \beta^{\text{T}})\text{G}$). For example, the sign of diagonal terms of the curvature matrix dictate whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) are indicative of selection for trait integration (less trait independence). Therefore, we can infer the indirect effects of selection on genetic constraints by counting the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix.

@@

Rather than imposing selection, parasitoids may themselves influence the expression of herbivore traits which could bias measurements of selection. In our system, it was plausible that parasitoids may influence chamber diameter by altering larval feeding behavior or killing larvae before they complete their development. To estimate this potential bias, we subset our data to only include galls where there was variation in larval survival within the same gall (i.e. 1 > mean survival > 0). If we assume that larvae within each gall should have similar chamber diameters because they come from the same clutch and experience the same local environment (an assumption our data supports: gall ID explains 54% of the variance in chamber diameter), then the relationship between chamber diameter and larval survival in this data subset represents the effect of parasitism on trait expression (i.e. bias). We used a GLMM with the same structure as described above except that we only modelled a linear relationship between chamber diameter and larval survival ($\alpha_{\text{Diam}}$). We detected a positive bias in both food-web treatments (complex $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebComplex:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.high,2)`]; simple $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebSimple:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.high,2)`]), indicating that unadjusted relationships would overestimate the strength of selection on chamber diameter. To account for this bias, we substracted our mean estimates of bias from our estimates with the full dataset prior to calculating chamber diameter's trait-fitness relationship and directional selection gradient.

@@

The structure of these matrices indicate that there are more selective constraints imposed on the gall midge in the simple (6 of 6) vs. complex (1 of 6) food web. Simple food webs acted to decrease genetic variance for all three phenotypic traits (negative diagonal terms), whereas only one trait (chamber diameter) experienced a decrease in additive genetic variance in the complex food web. For genetic covariances, the simple food web favored integration among all three phenotypic traits (nonzero off-diagonal terms), and thus constraints along all three axes of covariance (fig. \ref{fig:MV_Landscape}). In contrast, there was no strong evidence of trait integration in the complex food web (fig. \ref{fig:MV_Landscape}).

@@

>>>>

**Selection surface**: Since larval survival was our measure of absolute fitness, we used a GLMM that assumed a binomial error distribution (and logit-link function). To approximate the selection surface, we modelled larval survival as a function of food-web treatment as well as linear ($\alpha_{z_i}$), quadratic ($\alpha_{z_i:z_i}$), and linear interactions ($\alpha_{z_i:z_j}$) between each trait. We also allowed these trait-fitness relationships ($\alpha$) to vary between food-web treatments. Note that to obtain valid estimates of linear trait-fitness relationships, we removed nonlinear terms prior to estimating linear relationships [@Lande1983]. Other approaches have been advocated for approximating selection surfaces [@Schluter1988]; however, our approach enables us to calculate selection gradients, and thus is more appropriate for approximating the adaptive landscape [@Arnold2003]. To account for the nonindependence of clutch size (measured at gall level) and oviposition preference (measured at plant level) as well as any independent effects of willow genotype on larval survival, we modelled gall ID nested within plant ID nested within genotype ID as random effects. Although statistical models with random effects are not common in analyses of natural selection, we think that modelling random effects can mitigate biased estimates of selection due to environmental covariances between traits and fitness [@Rausher1992]. Since our end goal was to characterize the relationship between mean trait values and mean fitness (adaptive landscape), we assumed the mean value of our random effects (i.e., setting them to zero) when calculating trait-fitness relationships. We then used parametric bootstrapping (1,000 replicates) to estimate the effect of food-web treatment on larval survival as well as trait-fitness relationships in each food-web treatment. To determine whether trait-fitness relationships differed between food-web treatments, we calculated the difference in bootstrapped replicates between treatments.

@@

**Evolutionary constraints**: We quantified the effects of food-web complexity on evolutionary constraints by inspecting the slope and curvature of the adaptive landscape. <!--Adaptive landscape**: We took advantage of existing theory that translates estimates of selection gradients to the adaptive landscape [@Phillips1989; @Arnold2003]. The shape of the adaptive landscape can be described by both its slope and curvature, each of which describe different aspects of constraints on phenotypic evolution. Specifically, -->The number of selective constraints is determined by the slope of the adaptive landscape<!-- corresponds to the column vector of directional selection gradients-->, which in our study corresponds to:

@@

$$\text{Slope} = \beta = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} $$
<!--This slope describes the selective constraints acting on the trajectory of phenotypic evolution, which is governed by the 'Lande equation', $\Delta \bar z=\text{G}\beta$ [@Lande1979]. Quantitative predictions of evolutionary trajectories require knowledge of a population's **G**-matrix; however, We quantified the degree to which phenotypic evolution was constrained by selection <!--can gain qualitative insight to evolutionary constraints by counting the number of phenotypic traits under directional selection [selective constraints, @Arnold1992].-->
By comparing the number of traits that determine the slope (i.e., nonzero directional selection gradients) in simpler vs. more complex food webs, we can infer the effect of food-web complexity on selective constraints.

The indirect effects of selection on genetic constraints is governed by the curvature of the adaptive landscape, which in our study corresponds to:
<!--The curvature (**C**) of the adaptive landscape is a matrix that is determined by directional, quadratic, and correlational selection gradients, such that \boldmath$\text{C}=\gamma - \beta \beta^\text{T}$, or more explicitly:-->

@@

Note that we ommitted the upper triangle of each matrix for clarity since it is simply the reflection of the lower triangle. <!--This curvature matrix describes the indirect effects of selection on genetic constraints within a generation ($\Delta \text{G}=\text{G}(\gamma -\beta \beta^{\text{T}})\text{G}$). For example, -->The sign of diagonal terms of the curvature matrix dictate whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) are indicative of selection for trait integration (i.e., less independence). Therefore, we can infer the indirect effects of selection on genetic constraints by counting the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix.

@@

Rather than imposing selection, parasitoids may influence the expression of herbivore traits which could bias measurements of selection. In our system, it was plausible that parasitoids may influence chamber diameter by altering larval feeding behavior or killing larvae before they complete their development. To estimate this potential bias, we subset our data to only include galls where there was variation in larval survival within the same gall (i.e. 1 > mean survival > 0). If we assume that larvae within each gall should have similar chamber diameters because they come from the same clutch and experience the same local environment (an assumption our data supports: gall ID explains 54% of the variance in chamber diameter), then the relationship between chamber diameter and larval survival in this data subset represents the effect of parasitism on trait expression (i.e. bias). We used a GLMM with the same structure as described above except that we only modelled a linear relationship between chamber diameter and larval survival ($\alpha_{\text{Diam}}$). We detected a positive bias in both food-web treatments (complex $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebComplex:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.high,2)`]; simple $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebSimple:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.high,2)`]), indicating that unadjusted relationships would overestimate the strength of selection on chamber diameter. To account for this bias, we substracted our mean estimates of bias from our estimates with the full dataset prior to calculating chamber diameter's trait-fitness relationship and directional selection gradient.

@@

Remember that we can infer the indirect effects of selection on genetic constraints by counting the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix. The structure of these matrices indicates that there are more selective constraints imposed on the gall midge in the simpler (6 of 6) vs. more complex (1 of 6) food web. Simpler food webs acted to decrease genetic variance for all three phenotypic traits (negative diagonal terms), whereas only one trait (chamber diameter) experienced a decrease in additive genetic variance in the more complex food web. For genetic covariances, the simpler food web favored integration among all three phenotypic traits (nonzero off-diagonal terms), and thus constraints along all three axes of covariance (fig. \ref{fig:MV_Landscape}). In contrast, there was no strong evidence of trait integration in the more complex food web (fig. \ref{fig:MV_Landscape}).

@@

# Appendix
